DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-05-30 manuscript/manuscript.Rmd
edits to primary secondary calib

While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply  calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond]. One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications [@ksepka2015fossil]. This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
 <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->

In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record). Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical]. <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations. In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs. S2-S20). We also note that nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies. The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].

We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying]. It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s]; with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic]; and with methods that summarize age data unto a tree topology, such as DateLife.

>>>>

DateLife constructs summary chronograms using node ages extracted from existing chronograms, i.e. secondary calibrations.
In general, the scientific community has more confidence in chronograms using primary calibrations, where the dated tree is generated from a single analysis where carefully chosen fossil calibrations are the source of absolute time information, than in analyses dated using secondary calibrations [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
However, implementation of primary calibrations is difficult: it requires specialized expertise and training to discover, place and apply calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond]. One approach is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
The Fossil Calibration Database provides data for 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewed publications [@ksepka2015fossil]. This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
Yet, users still require the expertise to locate and calibrate appropriate nodes in their phylogenies which correspond with fossils available in the database.
 <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->

Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations using node dates based on previous molecular clock analyses can be as good as primary calibrations. 
Using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases [@shaul2002playing; @graur2004reading; @sauquet2013practical]. <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides insight into the application of secondary calibrations. Node ages summarized with DateLife and those from the original studies are well correlated (Supplementary Figs. S2-S20). We also note that DaetLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using our secondary calibrations than ages from the original studies. The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec]. However, these younger dates are likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the @burns2014phylogenetics tree were estimated using a secondary calibration based on a previously estimated molecular evolution rate across birds, and were major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (figure ??).

Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
